Glyph-maps for Visually Exploring Temporal Patterns in Climate Data and Models

### 2012-03-06 revision/paper.tex
tweaked abstract

The multivariate spatio-temporal nature of climate data makes it difficult to draw all of the aspects simultaneously. This paper describes the conceptualization and construction of a type of display, a glyph-map, that has been used infrequently for climate data. Glyph-maps are a specialization of multivariate glyph plots. Each spatial location is displayed with one glyph that represents the multiple measurements, recorded over time, at that location. Glyph-maps allow the discovery of both local and global structure, with a particular focus on temporal relationships. They provide alternatives to colored, facetted maps, or statistical summaries such as principal components. With the new conceptualization, developing interactive versions of glyph-maps will be easier, and it will be simpler to explore the perceptual effects of different scales. The methods are developed for rectangularly gridded data, but with some clever processing, which is explained, it is possible get good glyph-maps of irregularly gridded data. Guides, and reference marks, for different types of glyphs used in the glyph-maps are also discussed.

@@

\includegraphics[width=0.33\linewidth]{usa-lin-legend}
  %\includegraphics[width=0.25\linewidth]{ghcn-mountains}%
  \caption{(Top) Glyph-maps of the linear trend in ground station temperature values, 1950--2010, for USHCN stations, illustrating three issues for irregularly gridded data: overlapping icons, no natural choice for icon size, and the disordered appearance making comparisons more difficult. (Bottom) Legend showing
    temperature and temporal ranges of individual glyphs. }% (left) and stations in the mountain states (right).} 
  \label{fig:irregular}
\end{figure}
%The two panels illustrate one approach to the overlapping problem: zooming.  
One solution to the overlapping problem is to zoom in on smaller regions. There is an interplay between icon size (in geographical units), area plotted (in geographical units) and plot size (in display units).  Icons need to be big enough to be readable on the display, but small enough to avoid overlapping.  Readability can be maintained and overlapping reduced by decreasing the icon size and simultaneously increasing the display size or decreasing the area plotted.
% (as done in the right panel in Figure \ref{fig:irregular}).  
However, there are generally limitations in the maximum display size (or resolution) and the utility of plotting very small areas.

@@

\includegraphics[width=0.20\linewidth]{usa-season-legend}\hspace{2in}%
\includegraphics[width=0.20\linewidth]{usa-lin-legend}
  \caption{Glyph-maps of seasonal patterns (top left) and linear trends (top right) of USHCN stations, 1950--2010.  Nearby stations whose icons would overlap have been collapsed.  Each location has a glyph in the seasonal plot and color of the tiles corresponds to the average temperature of the locations.  Heavy lines in the linear trend plot correspond to average trends over more than one location. (Bottom) Legends showing scale of glyphs.} % Legend in Figure~\ref{fig:irregular-legend}}
  \label{fig:irregular-collapsed}
\end{figure}

@@

  \includegraphics[width=0.20\linewidth]{usa-season-legend}\hspace{2in}%
  \includegraphics[width=0.20\linewidth]{usa-lin-legend}
  \caption{Glyph-maps of seasonal patterns (top left) and linear trends (top right) of USHCN stations, 1950--2010.  Station locations have been rounded to the nearest degree Each location has a glyph in the seasonal plot and color of the tiles corresponds to the average temperature of the locations. Heavy lines in the linear trend plot correspond to average trends over more than one location. (Bottom) Legends showing scale of glyphs.} %Legend in Figure~\ref{fig:irregular-legend}}
  \label{fig:irregular-grid}
\end{figure}

>>>>

The multivariate spatio-temporal nature of climate data makes it difficult to draw all of the aspects simultaneously. This paper describes the conceptualization and construction of a type of display, a glyph-map, that can show these multiple aspects. Glyph-maps are a specialization of multivariate glyph plots. Each spatial location is displayed with one glyph that represents the multiple measurements, often recorded over time, at that location. Glyph-maps allow the discovery of both local and global structure, with a particular focus on temporal relationships, important for studying climate change. They provide alternatives to colored, facetted maps, or statistical summaries such as principal components. Glyph-maps have been used sporadically for spatio-temporal data, and with the ideas and software described in this paper it will be easier to produce them. The conceptualization described here also will enable developing interactive versions of glyph-maps, and make it simpler to explore the perceptual effects of different scales. The methods are developed for rectangular gridded data, but with some clever processing, which is explained, it is possible get good glyph-maps of irregularly gridded data. Guides and reference marks, for different types of glyphs used in the glyph-maps, are also discussed.

@@

\includegraphics[width=0.33\linewidth]{usa-lin-legend}
  %\includegraphics[width=0.25\linewidth]{ghcn-mountains}%
  \caption{(Top) Glyph-maps of the linear trend in ground station temperature values, 1950--2010, for USHCN stations, illustrating three issues for irregularly gridded data: overlapping icons, no natural choice for icon size, and the disordered appearance making comparisons more difficult. Local centering, subtracting the location mean from the linear fit, is used as the variation in average temperature is of much greater magnitude than the variation in linear trend. (Bottom) Legend showing
    temperature and temporal ranges of individual glyphs. The local scaling is apparent here with all lines going through zero at the mean year. }% (left) and stations in the mountain states (right).} 
  \label{fig:irregular}
\end{figure}
%The two panels illustrate one approach to the overlapping problem: zooming.  
One solution to the overlapping problem is to zoom in on smaller regions. There is an interplay between icon size (in geographical units), area plotted (in geographical units) and plot size (in display units).  Icons need to be big enough to be readable on the display, but small enough to avoid overlapping.  Readability can be maintained and overlapping reduced by decreasing the icon size and simultaneously increasing the display size or decreasing the area plotted.
% (as done in the right panel in Figure \ref{fig:irregular}).  
However, there are generally limitations in the maximum display size (or resolution) and the utility of plotting very small areas.

@@

\includegraphics[width=0.20\linewidth]{usa-season-legend}\hspace{2in}%
\includegraphics[width=0.20\linewidth]{usa-lin-collapse-legend}
  \caption{Glyph-maps of seasonal patterns (top left) and linear trends (top right) of USHCN stations, 1950--2010.  Nearby stations whose icons would overlap have been collapsed.  Each location has a glyph in the seasonal plot and color of the tiles corresponds to the average temperature of the locations.  Heavy lines in the linear trend plot correspond to average trends over more than one location. (Bottom) Legends showing scale of glyphs.} % Legend in Figure~\ref{fig:irregular-legend}}
  \label{fig:irregular-collapsed}
\end{figure}

@@

  \includegraphics[width=0.20\linewidth]{usa-season-legend}\hspace{2in}%
  \includegraphics[width=0.20\linewidth]{usa-lin-grid-legend}
  \caption{Glyph-maps of seasonal patterns (top left) and linear trends (top right) of USHCN stations, 1950--2010.  Station locations have been rounded to the nearest degree Each location has a glyph in the seasonal plot and color of the tiles corresponds to the average temperature of the locations. Heavy lines in the linear trend plot correspond to average trends over more than one location. (Bottom) Legends showing scale of glyphs.} %Legend in Figure~\ref{fig:irregular-legend}}
  \label{fig:irregular-grid}
\end{figure}
